Bayesian experimental design using regularized \\ determinantal point processes

### 2022-08-23 design/equivalents_nips/response.tex

\documentclass{article}

\usepackage{../sty/neurips2020/neurips_2020_author_response}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{verbatim}
\usepackage{lipsum}

\begin{document}

Two reviewers rated this paper very highly.
The other reviewer was anomalously low and provided a review that
lacked substance and was completely inappropriate.  
We thus ask the senior PC member to discount this review and make a
decision based on the other two thoughtful reviews, or to find
another reviewer to provide a more appropriate review.
(We will, however, address to the extent possible the comments of this reviewer below.)

\vspace{-3mm}

\paragraph{To Reviewer \#2} Thanks for the clear accept and the
helpful suggestions. In the final version, we will clarify the
discussion of $\gamma$ as a function of $k$ in Section~1.2, and we
will also expand Section~1.3 (note that a detailed discussion of
randomized iterative optimization is provided in Appendix B).

\vspace{-3mm}

\paragraph{To Reviewer \#3} Thanks for the clear accept and the helpful
suggestions. It is true that our proofs define a high probability
event, however this is merely for the analysis. The final
statements of Theorems 1 and 2 hold absolutely, rather than with high
probability. In the final version, we will expand our discussion of
low-rank approximation metrics.

\vspace{-3mm}

\paragraph{Addressing Reviewer \#4}
Below, we demonstrate that the review is inappropriate and should be
discounted, because:\vspace{-2mm}
\begin{enumerate}
\item The reviewer completely misrepresents and fails to understand
  the stated aims of the paper.
  \vspace{-1mm}
\item The reviewer makes sweeping claims which are technically wrong
  and unsupported.
\end{enumerate}\vspace{-2mm}

First of all, while we agree that the criticisms voiced by the
reviewer apply to many NeurIPS papers, this paper is \emph{not one of
  them}. We are well aware of the TCS work on low-rank
approximation and sketching. If we were to cite all of those TCS papers, it would
include hundreds of references (we choose to cite a few reviews, as is
common when an area gets to a certain level of maturity). However, more importantly,
unlike the TCS papers, in this work we are \emph{not} interested in obtaining
worst-case approximation or concentration bounds (see, e.g., lines 4
and 27). Yet, the reviewer appears to be
confused about this and states that ``the authors
show that, given a matrix A and a sketch S, the difference of the
singular values of $A^TA$  and $A^TS^TSA$ is concentrated around the
expectation.'' This is \emph{not at all} accurate in describing our
results. Instead, our goal is to provide a precise characterization,
which goes \emph{beyond worst-case bounds}, for the expected residual
projection matrix, $\mathbb{E}[I-(SA)^\dagger SA]$ (i.e., approximating an
analytically intractable deterministic quantity with a simpler
analytically tractable expression). Thus, in the context of low-rank
approximation, our goal is not
to improve on a TCS-style approximation objective (e.g., by showing a
$1+\epsilon$ error bound relative to the best rank $k$ subspace),
but rather to express the error in a simple form as a function of the
spectrum of the data matrix.  Also, unlike standard worst-case analysis, our analysis
does \emph{not} rely on satisfying some notion of the \emph{subspace
  embedding} property, which
significantly differentiates our work from that cited by the
reviewer. Note that a subspace embedding is
neither sufficient nor necessary for many numerical implementations of
sketching \citep{blendenpik,lsrn},
or statistical results
\citep{GarveshMahoney_JMLR,dobriban2019asymptotics,yang2020reduce},
as well as in the context of iterative optimization and
implicit regularization (see Section~1.3), which are discussed in
detail in the paper.

Finally, we point out the false and unsupported claims made by the
reviewer when comparing our paper to prior work.  This
likely arises from the reviewer's confusion regarding the nature of our results. (Meanwhile, the other two reviewers describe the paper as
``very well-written'' and ``clearly written''.)
Regarding \cite{optimal-matrix-product}, Reviewer \#4 states
that ``Main result here is a generalization of theorem 1'', and then
later, regarding the submission, the reviewer claims that ``the results presented here are, in
one form or another, either known results from TCS literature, or easy
corollaries''. The latter statement is incredibly broad, completely
unsupported and simply false, so we focus on the former. Regarding the former claim, Cohen et al. do provide a
low-rank approximation guarantee for sub-Gaussian sketches. However,
this result differs from ours in several respects. First of
all, instead of analyzing $\tilde A=A(SA)^\dagger SA$ directly as a
low-rank approximation with a sketch of size
$k$ (as we do), they use a larger sketch
(e.g., of size $C k/\epsilon^2$, where $C>1$ and $\epsilon< 1$) and
consider the matrix
$\tilde A _k$, defined as the best rank $k$ approximation of $\tilde
A$. This distinction is crucial for their analysis, which relies on
showing that a sketch of size sufficiently larger than $k$ ensures
a rank $k$ subspace embedding condition. This condition is not known to hold
(at least in the worst case) if the sketch is of size $k$. Our
novel analysis completely avoids subspace embeddings (which, as the reviewer points out, are central to TCS-style analysis).  This is why we can still provide
upper/lower bounds for the low-rank approximation error in this
important case. Also, the form of our bounds is completely different than that
of Cohen et al., in that we compare the error with a certain implicit
function of the singular values of $A$, which is different from the
error of the best rank $k$ approximation (used by Cohen et al.), and
so the role of $\epsilon$ in our paper is different than in
theirs. All in all, different methods are being analyzed, different types of
bounds are obtained, and completely different analysis is
used. Thus, by all indications, the reviewer is wrong, and our result is \emph{not} a corollary or a special case
of the results of Cohen et al.

\vspace{-.7\baselineskip}
\renewcommand\refname{\normalsize References}
{\scriptsize
\setlength{\bibsep}{0pt plus 0.2ex}
\bibliographystyle{abbrvnat}
\bibliography{../pap}
}

\end{document}

### 2022-08-23 design/nystrom-neurips2020-rebuttal.tex

\documentclass{article}

\usepackage{sty/neurips2020/neurips_2020_author_response}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{verbatim}
\usepackage{lipsum}
\usepackage{amsmath}
\begin{document}
We thank all the reviewers for the positive comments.

\paragraph{R1: Correctness of sr(A)-1 < k < sr(A)} 
Yes, this statement is correct. Our goal in Remark 1 is to illustrate a sharp transition in the approximation factor as $k$ approaches the stable rank from below. Remark 1(2) is an existential result i.e. there exists an $\mathbf{A}$ for which the lower bound on the approximation factor holds. 
\vspace{-2mm}
\paragraph{R1: L92 ``stable rank of A is 20" should be 40}
Here, we are referring to $\text{sr}_0(\mathbf{A})$, which is 20 and in Figure
1 it coincides with the first peak in the approximation factor.
\vspace{-2mm}
\paragraph{R1: Minor typos/comments}
Thanks for the careful read. We will make the suggested changes in the final version.
\vspace{-2mm}
\paragraph{R1: Fig 2.  Is the min $\Phi_s(k)$ line taken over $s\in\{10, 15, 25\}$ or over all values of $s\in [0,\text{rank}(\mathbf{A})]$ } The minimum is taken over all values of $s$. 
\vspace{-2mm}
\paragraph{R1: Proof of Thm 3. Importance of orthogonality of simplices}
For our setup, because of multiple simplices, the symmetry that holds for~\citet{pca-volume-sampling} breaks, and thus different subsets of the same size have different possible errors. Orthogonality, along with the choice of $\alpha_i, \beta_i$, ensures that we can pinpoint the best possible set of size $k$ and its corresponding error amongst the different choices (see proof of Lemma 4 in Appendix F). 
\vspace{-2mm}
\paragraph{R1: Other works on beyond worst case analysis}
We feel that any thorough discussion of works on ``beyond worst-case
analysis'' would have to span a very wide range of topics, which is
beyond the scope of this paper. In the final version, we will
reference \cite{roughgarden2019beyond} which provides a broader discussion on the area.
% In interest of keeping the paper precise, we did not feel that discussion of works for beyond worst case analysis on unrelated problems added anything of value to the exposition.
If the reviewer has any other works in mind which are particularly
relevant, we will definitely consider adding them.

\vspace{5mm}

\paragraph{R2: Suggested additional related works on matrix approximation}
The list of published papers on algorithms for matrix
approximation is indeed extensive. We thank the reviewer for the additional
pointers, as they are relevant to our work. In particular, we
believe that extending our results to the $\ell_p$ low-rank approximation
setting is an interesting direction for future work.

\vspace{5mm}

\paragraph{R3: Connections with overparametrization}
While we are not aware of a formal connection between the
multiple-descent phenomenon in column subset selection and
overparameterization in machine learning, both settings have to do
with obtaining a stable approximation of the data, and this stability can often be
captured by the spectral properties of the problem.
% We have established that the multiple descent phenomenon we observe is
% dependent on the spectrum of the matrix. Beyond that, there is no
% visible relationship of our work to overparameterization.

\vspace{5mm}

\paragraph{R4: Difference between k-DPP and greedy on enuite2001}
We also found the different behaviors of the approximation factor for
the two methods intriguing. One explanation for this is that the
expected approximation factor for a k-DPP is fully determined by the
eigenvalues, whereas the performance of the greedy method depends on
other characteristics of the data. However, it is not clear to us why
this is most pronounced on the eunite2001 dataset.
\vspace{-2mm}
\paragraph{R4: Correcting the effect of noisy eigenvalues}
We have not considered this, but it is indeed an
interesting question. One simple observation is that if one chooses the subset
size $k$ small enough so that the noisy tail of the spectrum consists of
only the eigenvalues with index sufficiently larger than $k$, then the effect of noise
on the approximation factor should be~minimal.

{
\setlength{\bibsep}{0pt plus 0.2ex}
\bibliographystyle{abbrvnat}
\bibliography{pap}
}
\end{document}

### 2022-08-23 design/aistats/bayesian.tex
Commit original repos

\newif\ifisarxiv
\isarxivtrue

% ready for submission
\ifisarxiv
  \documentclass[11pt]{article}
  \usepackage{fullpage}

  \usepackage{natbib}
\else
  \documentclass[twoside]{article}
  \usepackage{aistats2020}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{bm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{xfrac}
\usepackage{microtype}      % microtypography
\usepackage{color}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}

\input{shortdefs}
\newcommand{\mub}{\bm{\mu}}

@@

\begin{document}
\ifisarxiv
  \title{Bayesian experimental design using regularized \\
    determinantal point processes}
  \author{%
          Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
  Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \And
   Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
    % examples of more authors
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \AND
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
  }
\else
  \twocolumn[
  \aistatstitle{Bayesian experimental design \\using regularized 
    determinantal point processes}
\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3
}
  % \aistatsauthor{%
  %   Micha{\l } Derezi\'{n}ski
  %   \And
  %   Feynman Liang
  %   \And
  %   Michael W. Mahoney
  % }
  % \aistatsaddress {
  %   Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{mderezin@berkeley.edu}
  %   \And
  %   Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{feynman@berkeley.edu}
  %   \And
  %   ICSI and Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{mmahoney@stat.berkeley.edu}
  % } 
  ]
\fi

@@

Consider a collection of $n$ experiments parameterized by
$d$-dimensional vectors $\x_1,\dots,\x_n$, and let $\X$ denote the
$n\times d$ matrix with rows $\x_i^\top$. The outcome of the $i$th
experiment is a random variable $y_i = \x_i^\top\w + \xi_i$, where
$\w$ is the parameter vector of a linear model with prior distribution
$\Nc(\zero,\sigma^2\A^{-1})$, and $\xi_i\sim \Nc(0,\sigma^2)$ is
independent noise. In experimental design, we have access to the vectors 
$\x_i^\top$, for $i\in\{1,\ldots,n\} = [n]$, but we are allowed to
observe only a small number of outcomes $y_i$ for experiments we choose.
Suppose that we observe the outcomes from a subset
$S\subseteq [n]$ of $\lvert S \vert = k$ experiments. The
posterior distribution of $\w$ given $\y_S$ (the vector of outcomes in $S$) is:
\begin{align*}
  \w\mid \y_S \ &\sim\ \Nc(\mub, \Sigmab), \\
  \text{where } \mub &= (\X_S^\top\X_S + \A)^{-1}\X_S^\top\y_S, \\
  \Sigmab &= \sigma^2(\X_S^\top\X_S+\A)^{-1}.
\end{align*}
Here $\X_S$ is the $k\times d$ matrix with rows $\x_i^\top$ for
$i\in S$.

In Bayesian experimental design \citep{bayesian-design-review},
the prior precision matrix $\A$ is used to encode prior knowledge
and our goal is to choose $S$ so as to 
  minimize a function (a.k.a.~an optimality criterion) measuring the ``size''
  of the posterior covariance matrix
  $\Sigmab_{\w\mid\y_S} = \sigma^2(\X_S^\top\X_S+\A)^{-1}$.
Note that $\Sigmab_{\w\mid\y_S}$ is well defined even if $\A$ is not
invertible (i.e., an ``improper prior''). In particular, it includes classical
experimental design as the special case $\A = \zero$, as well as
the ridge-regularized case for $\A=\lambda\I$.
Denoting $\Sigmab$ as the subset covariance $\X_S^\top\X_S$, we will
use $f_{\A}(\Sigmab)$ to represent the following standard Bayesian 
optimality criteria \citep{bayesian-design-review,optimal-design-pukelsheim}:
\begin{enumerate}
\item A-optimality:\quad $f_{\A}(\Sigmab) = \tr\big((\Sigmab+\A)^{-1}\big)$;
\item C-optimality:\quad \!\!$f_{\A}(\Sigmab) = \cb^\top(\Sigmab+\A)^{-1}\cb$ for
$\cb \in \R^d$;
\item D-optimality:\quad $f_{\A}(\Sigmab) = \det(\Sigmab+\A)^{-1/d}$;
  \item V-optimality:\quad $f_{\A}(\Sigmab) =
    \frac1n\tr\big(\X(\Sigmab+\A)^{-1}\X^\top\big)$.
  \end{enumerate}
  Applications including clinical trials
  \citep{ryan2015fully,ding2008bayesian,spiegelhalter2004incorporating,berry2002adaptive,stangl1998bayesian,flournoy1993clinical},
  medical imaging \citep{owen2016optimisation},
  biological process models \citep{ryan2016optimal},
  and materials science
  \citep{frazier2016bayesian,ueno2016combo,terejanu2012bayesian}
  all use these optimality criteria and thus stand to benefit from
  our contributions.
  % Our algorithm can also be directly applied to improve
  % approximate design methods which leverage normal-based approximations
  % \citep{overstall2018approach}.

@@

%\noindent
\textbf{Bayesian experimental design.}
Given an $n\times d$ matrix $\X$,
a criterion $f_{\A}(\cdot)$ and~$k\in[n]$,
efficiently compute or approximate
\begin{align*}
\argmin_{S \subseteq [n]} f_{\A}(\X_S^\top\X_S)\quad\text{subject
  to}\quad |S|=k.
\end{align*}
We denote the value at the optimal solution as $\opt$.
The prior work around this problem can be grouped into two research questions.
% categories, each asking a question about the optimum of the above
% problem, which we will denote $\opt =
% \min_{S:|S|=k}f_{\A}(\X_S^\top\X_S)$.
The first question asks when does there exist a
polynomial time algorithm for finding a $(1+\epsilon$)-approximation
for $\opt$.
The second question asks what we can
infer about $\opt$ just from the spectral
information about the problem, which is contained in the data covariance matrix
$\Sigmab_\X=\X^\top\X\in\R^{d\times d}$.

@@

\begin{figure*}[htbp]
  \vspace{-4mm}
  \centering
  \renewcommand{\arraystretch}{1.5}
\begin{tabular}{r||c|c|l|l}
  &Criteria&Bayesian& $k$ %& dep.~on $d$ & dep.~on $1/\epsilon$
  & Cost after SDP\\%of rounding\\
  \hline\hline
\small  \cite{tractable-experimental-design}
  &\small A,V&\xmark  &$d^2/\epsilon$
  % &\Red{quadratic}& \Orange{linear}
  &$n^2\cdot d$\\
\small  \cite{near-optimal-design}
  &\small A,C,D,E,G,V&\cmark
&$ d/\epsilon^2$
% &\Orange{linear}& \Red{quadratic}
  &$n\cdot kd^2$\\ 
\small  \cite{proportional-volume-sampling}
  &\small A,D&\xmark &$d/\epsilon$
  % & \Orange{linear}& \Orange{linear}
  & $n^4\cdot k^2d$\\ 
  \hline
  \small\textbf{this paper}&\small A,C,D,V&\cmark
    & $d_{\A}/\epsilon$
    % &\Green{none}& \Orange{linear}
  & $n\cdot kd + k^2d^2$
\end{tabular}
\vspace{2mm}
\captionof{table}{Comparison of SDP-based $(1+\epsilon)$-approximation algorithms for
  classical and Bayesian experimental design (X-mark means that only
  the classical setting applies).
In the cost analysis, $n$ could be replaced by the number of
non-zero weights in the SDP solution. For simplicity we omit the log
terms and assume that $\epsilon = \Omega(\frac1{d_{\A}})$. Our
approach beats other methods both in terms of the
runtime and the dependence of $k$ on $d$ (when $d_\A=o(d)$).} 
\label{tab:comp}
\vspace{-3mm}
\end{figure*}

@@

We now define our regularized DPP and describe its
connection with correlation and ensemble~DPPs.
\begin{definition}\label{d:r-dpp}
Given matrix $\X\in\R^{n\times d}$, a sequence  $p=(p_1,\dots,p_n)\in[0,1]^n$
and a psd matrix $\A\in\R^{d\times d}$ such that
$\sum_ip_i\x_i\x_i^\top\!+\A$ is
full rank, let
$\DPPreg{p}(\X,\A)$ be a distribution over $S\subseteq [n]$:
  \begin{align}
  \Pr(S) = \frac{\det(
  \X_S^\top\X_S+\A)}{\det\!\big(
 \sum_ip_i\x_i\x_i^\top+ \A\big)}
\cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in S}(1-p_i).\label{eq:poisson-prob}
\end{align}
%   \Pr(\pi)\ \propto\ \det\!\big(\A +
%   \X_{\pi}^\top\X_{\pi}\big)
%   \Pr\!\bigg(\Big[K\!=\!|\pi|\Big]\,\wedge\,\Big[\forall_{i=1}^{|\pi|}\,\pi_i\!=\!\pi_i\Big] \bigg).
% \end{align*}
\end{definition}
The fact that this is a proper distribution (i.e., that it sums to
one) can be restated as a determinantal expectation formula: if
$b_i\sim\mathrm{Bernoulli}(p_i)$ are independent Bernoulli random
variables, then
\begin{align*}
  &\sum_{S\subseteq [n]}\!\det(
  \X_S^\top\X_S+\A)\prod_{i\in S}p_{i}\prod_{i\not\in S}(1-p_i) \\
  &=
  \E\bigg[\!\det\!\Big(\sum_ib_i\x_i\x_i^\top\!+\A\Big)\bigg]\overset{(*)}{=}
  \det\!\Big(\sum_i\E[b_i]\x_i\x_i^\top\!+\A\Big),
\end{align*}
where $(*)$ was shown by \cite[Lemma
7]{determinantal-averaging} and $\E[b_i]=p_i$.
% follows from a recently shown determinantal expectation formula.
% \begin{lemma}[\cite{determinantal-averaging}]\label{l:cb}
% For $\X\!\in\!\R^{n\times d}$, $\A\!\in\!\R^{d\times d}$
% and a random set $S\subseteq [n]$ drawn with independent Bernoullis,
% \begin{align*}
%   \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)\Big]
%   = \det\!\Big( \E\big[\X_S^\top\X_S\big]+\A\Big).
% \end{align*}
% \end{lemma}

@@

    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1},\label{eq:sqinv}

@@

    \sum_ip_i\x_i\x_i^\top +\A \Big)^{-1}.\label{eq:det}

@@

       &=

@@

% \subsubsection*{Acknowledgements}

% MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
%   support of this work. 
%   Also, MWM and MD \Red{
%     thank the NSF TRIPODS program
%     and FL thanks the NPSC for funding.
%   }
%   The authors thank Uthaipon
%   T.~Tantipongpipat for valuable discussions.

\bibliographystyle{apalike}
\bibliography{pap}

>>>>

\newif\ifisarxiv
\isarxivfalse

% ready for submission
\ifisarxiv
  \documentclass[11pt]{article}
  \usepackage{fullpage}
\else
  \documentclass[twoside]{article}
  \usepackage[accepted]{aistats2020}

  % \special{papersize = 8.5in, 11in}
  % \setlength{\pdfpageheight}{11in}
  % \setlength{\pdfpagewidth}{8.5in}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{bm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{xfrac}
\usepackage{microtype}      % microtypography
\usepackage{color}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}

\input{../shortdefs}
\newcommand{\mub}{\bm{\mu}}

@@

\begin{document}
\ifisarxiv
  \title{Bayesian experimental design using regularized \\
    determinantal point processes}
  \author{%
          Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
  Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \And
   Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
    % examples of more authors
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \AND
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
  }
  \else

  \runningtitle{Bayesian experimental design using regularized DPPs}
  \twocolumn[
  \aistatstitle{Bayesian experimental design \\using regularized 
    determinantal point processes}
  \aistatsauthor{%
    Micha{\l } Derezi\'{n}ski
    \And
    Feynman Liang
    \And
    Michael W. Mahoney
  }
  \aistatsaddress {
    Department of Statistics\\
    University of California, Berkeley\\
    \texttt{mderezin@berkeley.edu}
    \And
    Department of Statistics\\
    University of California, Berkeley\\
    \texttt{feynman@berkeley.edu}
    \And
    ICSI and Department of Statistics\\
    University of California, Berkeley\\
    \texttt{mmahoney@stat.berkeley.edu}
  } 
% \aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
% \aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }
  ]
\fi

@@

Consider a collection of $n$ experiments parameterized by
$d$-dimensional vectors $\x_1,\dots,\x_n$, and let $\X$ denote the
$n\times d$ matrix with rows $\x_i^\top$. The outcome of the $i$th
experiment is a random variable $y_i = \x_i^\top\w + \xi_i$, where
$\w$ is the parameter vector of a linear model with prior distribution
$\Nc(\zero,\sigma^2\A^{-1})$, and $\xi_i\sim \Nc(0,\sigma^2)$ is
independent noise. In experimental design, we have access to the vectors 
$\x_i^\top$, for $i\in\{1,\ldots,n\} = [n]$, but we are allowed to
observe only a small number of outcomes $y_i$ for experiments we choose.
Suppose that we observe the outcomes from a subset
$S\subseteq [n]$ of $\lvert S \vert = k$ experiments. The
posterior distribution of $\w$ given $\y_S$ (the vector of outcomes in $S$) is:
\begin{align*}
  \w\mid \y_S \ &\sim\ \Nc(\mub, \Sigmab), \\
  \text{where } \mub &= (\X_S^\top\X_S + \A)^{-1}\X_S^\top\y_S, \\
  \Sigmab &= \sigma^2(\X_S^\top\X_S+\A)^{-1}.
\end{align*}
Here, $\X_S$ is the $k\times d$ matrix with rows $\x_i^\top$ for
$i\in S$.

In Bayesian experimental design \citep{bayesian-design-review},
the prior precision matrix $\A$ is used to encode prior knowledge
and our goal is to choose $S$ so as to 
  minimize a function (a.k.a.~an optimality criterion) measuring the ``size''
  of the posterior covariance matrix
  $\Sigmab_{\w\mid\y_S} = \sigma^2(\X_S^\top\X_S+\A)^{-1}$.
Note that $\Sigmab_{\w\mid\y_S}$ is well defined even if $\A$ is not
invertible (i.e., an ``improper prior''). In particular, it includes classical
experimental design as the special case $\A = \zero$, as well as
the ridge-regularized case for $\A=\lambda\I$.
Denoting $\Sigmab$ as the subset covariance $\X_S^\top\X_S$, we will
use $f_{\A}(\Sigmab)$ to represent the following standard Bayesian 
optimality criteria \citep{bayesian-design-review,optimal-design-pukelsheim}:
\begin{enumerate}
\item A-optimality:\quad $f_{\A}(\Sigmab) = \tr\big((\Sigmab+\A)^{-1}\big)$;
\item C-optimality:\quad \!\!$f_{\A}(\Sigmab) = \cb^\top(\Sigmab+\A)^{-1}\cb$ for
$\cb \in \R^d$;
\item D-optimality:\quad $f_{\A}(\Sigmab) = \det(\Sigmab+\A)^{-1/d}$;
  \item V-optimality:\quad $f_{\A}(\Sigmab) =
    \frac1n\tr\big(\X(\Sigmab+\A)^{-1}\X^\top\big)$.
  \end{enumerate}
  Applications including clinical trials
  \citep{ryan2015fully,ding2008bayesian,spiegelhalter2004incorporating,berry2002adaptive,stangl1998bayesian,flournoy1993clinical},
  medical imaging \citep{owen2016optimisation},
  materials science
  \citep{frazier2016bayesian,ueno2016combo,terejanu2012bayesian},
 and biological process models \citep{ryan2016optimal}
  all use these optimality criteria and thus stand to benefit from
  our contributions.
  % Our algorithm can also be directly applied to improve
  % approximate design methods which leverage normal-based approximations
  % \citep{overstall2018approach}.

@@

%\noindent
\textbf{Bayesian experimental design.}
Given an $n\times d$ matrix $\X$,
a criterion $f_{\A}(\cdot)$ and~$k\in[n]$,
efficiently compute or approximate
\begin{align*}
\argmin_{S \subseteq [n]} f_{\A}(\X_S^\top\X_S)\quad\text{subject
  to}\quad |S|=k.
\end{align*}
We denote the value at the optimal solution as $\opt$.
The prior work around this problem can be grouped into two research questions.
% categories, each asking a question about the optimum of the above
% problem, which we will denote $\opt =
% \min_{S:|S|=k}f_{\A}(\X_S^\top\X_S)$.
The first question asks when does there exist a
polynomial time algorithm for finding a $(1+\epsilon$)-approximation
for $\opt$.
The second question asks what we can
infer about $\opt$ just from the spectral
information about the problem, which is contained in the
data covariance matrix~$\Sigmab_\X=\X^\top\X$.

@@

\begin{figure*}[htbp]
  \vspace{-2mm}
  \centering
  \renewcommand{\arraystretch}{1.5}
\begin{tabular}{r||c|c|l|l}
  &Criteria&Bayesian& $k$ %& dep.~on $d$ & dep.~on $1/\epsilon$
  & Cost after SDP\\%of rounding\\
  \hline\hline
\small  \cite{tractable-experimental-design}
  &\small A,V&\xmark  &$d^2/\epsilon$
  % &\Red{quadratic}& \Orange{linear}
  &$n^2\cdot d$\\
\small  \cite{near-optimal-design}
  &\small A,C,D,E,G,V&\cmark
&$ d/\epsilon^2$
% &\Orange{linear}& \Red{quadratic}
  &$n\cdot kd^2$\\ 
\small  \cite{proportional-volume-sampling}
  &\small A,D&\xmark &$d/\epsilon$
  % & \Orange{linear}& \Orange{linear}
  & $n^4\cdot k^2d$\\ 
  \hline
  \small\textbf{this paper}&\small A,C,D,V&\cmark
    & $d_{\A}/\epsilon$
    % &\Green{none}& \Orange{linear}
  & $n\cdot kd + k^2d^2$
\end{tabular}
%\vspace{2mm}
\captionof{table}{Comparison of SDP-based $(1+\epsilon)$-approximation algorithms for
  classical and Bayesian experimental design (X-mark means that only
  the classical setting applies).
In the cost analysis, $n$ could be replaced by the number of
non-zero weights in the SDP solution. For simplicity we omit the log
terms and assume that $\epsilon = \Omega(\frac1{d_{\A}})$. Our
approach beats other methods both in terms of the
runtime and the dependence of $k$ on $d$ (when $d_\A=o(d)$).} 
\label{tab:comp}
\vspace{-3mm}
\end{figure*}

@@

We now define our regularized DPP and describe its
connection with correlation and ensemble~DPPs.
\begin{definition}\label{d:r-dpp}
Given matrix $\X\in\R^{n\times d}$, a sequence  $p=(p_1,\dots,p_n)\in[0,1]^n$
and a psd matrix $\A\in\R^{d\times d}$ such that
$\sum_ip_i\x_i\x_i^\top\!+\A$ is
full rank, let
$\DPPreg{p}(\X,\A)$ be a distribution over $S\subseteq [n]$:
  \begin{align}
  \Pr(S) = \frac{\det(
  \X_S^\top\X_S+\A)}{\det\!\big(\!
 \sum_ip_i\x_i\x_i^\top\!+ \A\big)}
\cdot \!\prod_{i\in S}p_{i}\cdot\!\prod_{i\not\in S}(1\!-\!p_i).\label{eq:poisson-prob}
\end{align}
%   \Pr(\pi)\ \propto\ \det\!\big(\A +
%   \X_{\pi}^\top\X_{\pi}\big)
%   \Pr\!\bigg(\Big[K\!=\!|\pi|\Big]\,\wedge\,\Big[\forall_{i=1}^{|\pi|}\,\pi_i\!=\!\pi_i\Big] \bigg).
% \end{align*}
\end{definition}
The fact that this is a proper distribution (i.e., that it sums to
one) can be restated as a determinantal expectation formula: if
$b_i\sim\mathrm{Bernoulli}(p_i)$ are independent Bernoulli random
variables, then
\begin{align*}
  &\sum_{S\subseteq [n]}\!\det(
  \X_S^\top\X_S+\A)\prod_{i\in S}p_{i}\prod_{i\not\in S}(1-p_i) \\
  &\!=
  \E\bigg[\!\det\!\Big(\sum_ib_i\x_i\x_i^\top\!+\A\Big)\bigg]\overset{(*)}{=}
  \det\!\Big(\sum_i\E[b_i]\x_i\x_i^\top\!+\A\Big),
\end{align*}
where $(*)$ follows  from Lemma 7 of \citet{determinantal-averaging}.
% follows from a recently shown determinantal expectation formula.
% \begin{lemma}[\cite{determinantal-averaging}]\label{l:cb}
% For $\X\!\in\!\R^{n\times d}$, $\A\!\in\!\R^{d\times d}$
% and a random set $S\subseteq [n]$ drawn with independent Bernoullis,
% \begin{align*}
%   \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)\Big]
%   = \det\!\Big( \E\big[\X_S^\top\X_S\big]+\A\Big).
% \end{align*}
% \end{lemma}

@@

  &\preceq \Big(\!
    \sum_ip_i\x_i\x_i^\top \!+\A \Big)^{-1}\!\!,\label{eq:sqinv}

@@

  &\leq\det\!\Big(\!
    \sum_ip_i\x_i\x_i^\top\!+\A \Big)^{\!-1}\!\!.\!\!\label{eq:det}

@@

@@

\subsubsection*{Acknowledgements}

MWM would like to acknowledge ARO, DARPA, NSF and ONR for providing partial
  support of this work. Also, MWM and MD thank the NSF TRIPODS program
    and FL thanks the NPSC for funding.  The authors thank Uthaipon
  T.~Tantipongpipat for valuable discussions.

\bibliographystyle{apalike}
\bibliography{../pap}
